Quantum-assisted unsupervised data clustering

### 2020-05-06 manuscript.tex
multiline style
fix latex style

At this stage, the information regarding the differences between pairs of $\{X\}$ and $\{Y\}$ \hl{encoded in the amplitudes, in order to extract the Hamming distances between the relevant $\left| x_i \right\rangle$, $\left| y_j \right\rangle$ we return to our initial basis} for register $\left| X \right\rangle$ by applying pairwise CNOT gates:
%
\begin{multline}
    \left| \psi_f \right\rangle =  
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
	\left[ 
	    \cos\left(
	        \dfrac{\pi}{2n}
	        \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij} 
	    \right)
        \left| x_i \right\rangle 
        \left| y_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+
        \left. i \sin\left(
            \dfrac{\pi}{2n}
            \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij} 
        \right)
        \left| x_i \right\rangle 
        \left| y_j \right\rangle 
        \left| 1 \right\rangle
    \right] .
\end{multline}
%
This makes $\{X\}$ store the input vectors again, as in the initial step 
and preserves the amplitudes of the auxiliary qubit which are proportional to how different each pairs of $\{X\}$ and $\{Y\}$ are.

We thus have the Hamming distances encoded into the amplitudes of the final state. 
From the statistics of the measurement outcomes, the distance matrix between two data sets of binary vectors can be obtained (Fig.~\ref{fig:distance_matrix}). 
In this case, the biggest amplitude of the measurement result coincides with the smallest Hamming distance when the measurement result of the ancilla qubit is 0. 
If the ancilla qubit is 1, the smallest amplitude of the measurement result coincides with the smallest Hamming distance.

@@

Currently available quantum platforms are still subject to substantial level of noise and extracting the exact distance from amplitude is still a difficult task. However, for most of algorithms, SOFM included, the information of nearest vectors is sufficient. Because of that property, those algorithms fall into a special case category for which, as there is only one input vector considered, the ``Decoding'' stage (Fig.\ref{fig:qcircuit}) can be removed as measurement no longer needs to indicate for which input vector the distance has been measured. In this special case scenario the circuit depth complexity is matching with \cite{shuld2014}, but in general case for multiple input vectors the ``Decoding'' stage still needs to be included leading to larger circuit depth and less attractive complexity in terms of number of controlled gate operations. The number of controlled gate operations in this general case of multiple input vectors is matching the number of controlled gate operations in \cite{trugenberger2001}.

>>>>

At this stage, the information regarding the differences between pairs of $\{X\}$ and $\{Y\}$ \hl{encoded in the amplitudes, in order to extract the Hamming distances between the relevant $\left| x_i \right\rangle$, $\left| y_j \right\rangle$ we return to our initial basis} for register $\left| X \right\rangle$ by applying pairwise CNOT gates:
%
\begin{multline}
	\label{eq:final_state}
    \left| \psi_f \right\rangle =  
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
	\left[ 
	    \cos\left(
	        \dfrac{\pi}{2n}
	        \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij} 
	    \right)
        \left| x_i \right\rangle 
        \left| y_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+
        \left. i \sin\left(
            \dfrac{\pi}{2n}
            \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij} 
        \right)
        \left| x_i \right\rangle 
        \left| y_j \right\rangle 
        \left| 1 \right\rangle
    \right] .
\end{multline}
%
This makes $\{X\}$ store the input vectors again, as in the initial step 
and preserves the amplitudes of the auxiliary qubit which are proportional to how different each pairs of $\{X\}$ and $\{Y\}$ are.

From the statistics of the measurement outcomes of final state~(\ref{eq:final_state}) we recreate the amplitudes of ancilla qubit states. 
From those amplitudes estimations we are able to plot the distance matrix between two data sets of binary vectors.
The probability amplitude of the ancilla qubit outcomes captures the exact Hamming distance as the result of the preprocessing function.
There are two possible outcomes of measurement of the ancilla qubit, each has own probability amplitude and own interpretation of that amplitude. 
For instance, for the \left| 0 \right\rangle outcome, the larger the amplitude the smaller the Hamming distance, 
and for the \left| 1 \right\rangle outcome it is the other way around, magnitude of the amplitude of that outcome is proportional to the Hamming distance.

@@

In comparison with the circuits for Hamming distance calculations proposed previously \cite{trugenberger2001}, 
our method allows one to reduce the number of gates in the circuit (at least in $2n$ one-qubit NOT gates)
and reconstruct all distance matrix in parallel and can be implemented with high fidelity.
